---
title: "CA8"
author: "CHDNAT002 Natalie Chidikwindi & SGNLIH001 Lihle Sigenu"
format: pdf
editor: visual
---

```{r setup}
#| label: libraries
#| warning: false
#| include: false
#| echo: false

if (!requireNamespace("DataTidyRodoSTA2005S", quietly = TRUE)) {
if (!requireNamespace("remotes", quietly = TRUE)) {
utils::install.packages("remotes")
}
remotes::install_github("MiguelRodo/DataTidyRodoSTA2005S")
}
if (!requireNamespace("tibble", quietly = TRUE)) {
install.packages("tibble")
}
library(tibble) 
library(pls)
data("data_tidy_yield", package = "DataTidyRodoSTA2005S")
dat <- data_tidy_yield

#packages
library(xtable)
library(tinytex)
library(latexpdf)
```

```{r}
# We removed the first two columns as they referred to the farm_id and maize 
X <- dat[,-c(1,2)]
```

# Question 1

## a) Estimated regression coefficients for:

### i) maize yield \~ soil PH + dist_water

```{r, results='asis'}
Y1 <- as.matrix(dat[,2]) #Maize Yield
X1 <- as.matrix(cbind(1,X[,c(6,8)]))
beta_vec1 <- as.matrix(solve(t(X1)%*%X1)%*%t(X1)%*%Y1)
colnames(beta_vec1) <- c("Coeffecient estimates")
rownames(beta_vec1) <- c("Intercept", "SoilPH", "Dist_water")

print(xtable(beta_vec1, digits = 2, align = "lc",
       caption = "Estimates for regression coeffecients"),comment = F)
```

### ii) sunlight \~ soil PH + dist_water

```{r, results='asis'}
Y2 <- as.matrix(dat[,7]) # Sunlight
X2 <- X1
beta_vec2 <- as.matrix(solve(t(X2)%*%X2)%*%t(X2)%*%Y2)
colnames(beta_vec2) <- c("Coeffecient estimates")
rownames(beta_vec2) <- c("Intercept", "SoilPH", "Dist_water")

print(xtable(beta_vec2, digits = 2, align = "lc",
       caption = "Estimates for regression coeffecients"), comment = F)
```

### iii) maizeYield + Sunlight \~ soil ph + dist_water

```{r, results='asis'}
Y3 <- as.matrix(cbind(Y1,Y2))
X3 <- X1
beta_vec3 <- as.matrix(solve(t(X3)%*%X3)%*%t(X3)%*%Y3)
colnames(beta_vec3) <- c("MaizeYield estimates", "Sunlight estimates")
rownames(beta_vec3) <- c("Intercept", "SoilPH", "Dist_water")

print(xtable(beta_vec3, digits = 2, align = "lcc",
       caption = "Estimates for regression coeffecients"), comment = F)
```

## b) Confidence Interval for temperature \~ soil PH

```{r, results='asis'}
#finding the estimates for temperature against soil
Y4 <- as.matrix(dat[,4])
X4 <- as.matrix(X1[,c(1,2)])
C11 <- solve(t(X1) %*% X1)[2:3, 2:3]
beta_vec4 <- solve(t(X4)%*%X4)%*%t(X4)%*%Y4

#Confidence interval
s <- sum((Y4 - X4 %*% beta_vec4) ^ 2) / (nrow(Y4) - ncol(X4))
var_resp <- s*solve(t(X4)%*%X4)
sd_beta_soil <- sqrt(var_resp[2,2])


#find t assiciated 
t_beta_soil <- qt(0.975, nrow(Y4) - ncol(X4))

#CI
CI_soil <- t((beta_vec4[2] + c(-1, 1) * t_beta_soil * sd_beta_soil))
CI_soil <- as.matrix(CI_soil)|>
signif(2)
colnames(CI_soil)<- c("2.5%", "97.5%")
rownames(CI_soil) <- ("SoilPH")
# print(xtable(CI_soil, digits = 2, 
# caption = "95% confidence interval for SoilPH against Temperature",
#align = "lcc"), comment = F, include.rownames = TRUE)}
```

```{=latex}
\begin{table}[ht]
\centering
\begin{tabular}{lcc}
  \hline
 & 2.5\% & 97.5\% \\ 
  \hline
SoilPH & 0.79 & 2.30 \\ 
   \hline
\end{tabular}
\caption{95\% confidence interval for SoilPH against Temperature} 
\end{table}
```

The confidence interval above indicates that if an arbitrarily large number of samples was drawn from the same population at the data, 95% of the time the \beta estimate for SoilPH would lie within the interval mentioned above.

## c) F-test for the effect of soil PH and distance to water on maize yield.

The null and alternative hypotheses are as follows:

$$H_0: \beta_{SoilPH} = \beta_{distwater} = 0$$ $$H_1: \text{At least 1 } \beta_i \neq 0 $$ where $i$ = {SoilPH, dist_water}

```{r, results='asis'}
# extracts SoilPH and dist_water

res_beta_vec1 <- as.matrix(beta_vec1[2:3])

s <- sum((Y1 - X1 %*% beta_vec1) ^ 2) / (nrow(Y1) - ncol(X1))

# f-stat
f_stat <- t(res_beta_vec1) %*% solve(C11) %*% res_beta_vec1 / (2 * s)

#p-value
p_value <- pf(f_stat, length(res_beta_vec1), 
              nrow(Y1) - ncol(X1), lower.tail = FALSE)

#table formatting
results <- cbind(f_stat, p_value)
colnames(results) <- c("F-stat", "p-value")
print(xtable(results, digits = 2, 
             caption = "F-statistic and p-value for hypothesis test"),
      comment = F)
```

The F-stat is high, and the p-value is very close to 0. This means it is unlikely that SoilPH and distance to water (dist_water) do not have a have a statistical effect on MaizeYield. We reject the null hypothesis. At least one predictor, SoilPH or dist_water, has a statistically significant effect on MaizeYield on average.

# Question 2

Perform multiple regression, PCR and PLS on the data_tidy_mali dataset. The goal is to predict the quantity of cotton planted, using the other variables.

```{r loading of data}
#| warning: false
#| include: false
#| echo: false
if (!requireNamespace("remotes", quietly = TRUE)) {
install.packages("remotes")
}
remotes::install_github("MiguelRodo/DataTidy23RodoHonsMult@2024")
data("data_tidy_mali", package = "DataTidy23RodoHonsMult")

df <- data_tidy_mali
```

## PCR approach

```{r}
X <- data_tidy_mali[,-3]
Y <- data_tidy_mali$cotton

# Calculate the principal components of the (non-intercept) predictors
X <- scale(X)

eigen_x <- eigen(cor(X))

pc_loadings <- eigen_x$vectors
rownames(pc_loadings) <- paste0("V", 1:8)
colnames(pc_loadings) <- paste0("PC", 1:8)

eigen_vals <- as.matrix(t(eigen_x$values))
colnames(eigen_vals) <- paste0("PC", 1:8)
rownames(eigen_vals) <- c("eigenvalue")

prop_var <- numeric(8)
for (i in 1:8){
  prop_var[i] = (sum(eigen_x$values[i])/sum(eigen_x$values))*100
}
#plot of each PC
par(mar = c(4, 4, 2, 1))
plot(prop_var, type = "b",
     xlab = "PC",
     ylab = "variance explained (%)",
     main  = "Scree Plot showing the proportion of variation explained",
     cex.main = 1,
     xaxt = "n")
axis(side = 1, at = 1:8, labels = 1:8)


# sum(eigen_x$values[1:2])/sum(eigen_x$values) * 100 =  57.69518
# sum(eigen_x$values[1:4])/sum(eigen_x$values) * 100 = 82.16656
```

Based on the plot above as well as the proportion of variance explained by each additional PC, we have decided to use the first 4 principal components. Although the first 2 may be sufficient for visualizing the data, they only explain 57.70% of the variation in our data. By add $\text{PC}_3$ and $\text{PC}_4$ the amount of variation explained by out principal components will be 82.17% , which will ensure more accurate estimates for our regression coefficients.

The linear regression estimates are then as follows:

```{r, results='asis'}
#calculating scores
score_matrix <- cbind(1,X %*%pc_loadings[,1:4])
beta_vector <- solve(t(score_matrix)%*%score_matrix)%*%t(score_matrix)%*%Y
beta_vector <- as.matrix(beta_vector)
colnames(beta_vector) <- c("Coeffecient estimates")
rownames(beta_vector) <- c("Intercept", paste0("PC", 1:4))

print(xtable(beta_vector, digits = 2,
       caption = "Estimates for regression coeffecients using PCR"), 
      comment = F)
```

## Multiple Regression

Under multiple linear regression, we obtain the following coefficient estimates:

```{r results='asis'}
#| label: tab-mlr-coefs

# Create design matrix
dX <- cbind(1, X)

# Obtain beta estimates
betas <- solve(t(dX)%*%dX)%*%t(dX)%*%Y

# Predict using estimates 
yhats <- dX%*%betas

# Compute R^2
SST <- sum((Y - mean(Y))^2)
SSE <- sum((Y-yhats)^2)
SSR <- SST - SSE
R_2 <- SSR/SST

colnames(betas) <- c("Coefficient estimates")
rownames(betas) <- c("Intercept", colnames(X))

print(xtable(betas, digits = 2,
       caption = "Estimates for regression coeffecients using MLR"), 
      comment = F)
```

\newpage

## PLS Approach

```{r}
# Fit the PLS model with 10-fold (default) cross validation
set.seed(2026)
pls_model <- plsr(Y~X, validation="CV")
par(mfrow = c(1,2))# Plot the cross-validation RMSE and R^2
validationplot(pls_model)
abline(v = 1, col = "red")
validationplot(pls_model, val.type = "R2")
abline(v=1, col = "red")
R2_vals <- drop(R2(pls_model)$val)
max <- max(R2_vals)
```

Using cross validation, we note that the Root Mean Square Error is minimized when using 1 PLS component. This also results in the maximum $R^2$ value, where the model explains `r round(max,2)`% of the variation in quantity of cotton planted.

```{r}
pls_scores <- scores(pls_model)[, 1, drop = FALSE]

gamma_hat <- solve(t(pls_scores) %*% pls_scores) %*% t(pls_scores) %*% Y
```

And so the linear regression estimate for this PLS component is given by $\gamma =$ `r round(gamma_hat,2)`.

The table below provides a comparison of percentage of variation explained in quantity of cotton planted for each model.

```{r, results='asis'}
r2 <- matrix(c(82.17, round(R_2*100,2), round(max*100,2)))
colnames(r2) <- c("Proportion of Variance")
rownames(r2) <- c("PCR", "MLR", "PLS")

print(xtable(r2, digits = 2,
       caption = "Proportion of Variance Explained using PCR, MLR and PLS"), 
      comment = F)
```

From the table above, PCR performed best followed by multiple regression , and with PLS reporting the lowest explained variation. This suggests that the explanatory variables are moderately good at predicting the quantity of cotton planted but they suffer from multi-collinearity which is captured by PCR. Multiple Regression assumes zero multi-collinearity present and so assigns weightings accordingly, resulting in redundancy. PLS, in this case, performs the worst. PLS factors the response variable when creating components but if the true relationship is well explained by the original variables it may introdce noise rather than improving prediction.
